
\subsubsection{Expressions}

We show that for every expression~$e$ in the script,
\begin{eqnarray*}
f(\eval \rho ~ e) & = \eval (f \after \rho) e .
\end{eqnarray*}

\framebox{\ldots}

Note:
\begin{eqnarray*}
\eval \rho ~(g(e)) & = & (\eval \rho ~g) (\eval \rho~ e) \\
\eval \rho ~(\lambda x \spot e) & = & 
  \lambda z \spot \eval (\rho \oplus \set{x \mapsto z})~ e.
\end{eqnarray*}


Case $e = t$, i.e.~the name of the distinguished type ...


We now consider process expressions.

\paragraph{Basic processes.} $STOP$, $SKIP$, $div$: these are all trivial. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{proof4-prefix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Non-indexed sequential operators.}

Below we write $\extchoice$ for a semantic operator: ``${P \extchoice Q}$''
represents an external choice between~$P$ and~$Q$, each of which is an LTS\@.
We have
%
\begin{eqnarray*}
f(P \extchoice Q) & = & f(P) \extchoice f(Q).
\end{eqnarray*}
%
For every transition $P \trans{e} P'$, the initial state of each of the above
processes has a transition labelled $f(e)$ to~$f(P')$; and similarly for
transitions of~$Q$.

We then have the following (for syntactic expressions~$P$ and~$Q$):
%
\begin{calc}
& f(\eval \rho ~ (P \; \M{[]} Q) \\
= & f((\eval \rho~P) \extchoice (\eval \rho~Q)) \\
= & \com{above result} \\
& f(\eval \rho~P) \extchoice f(\eval \rho~Q) \\
= & \com{inductive hypothesis} \\
& (\eval (f\after\rho)~P) \extchoice (\eval (f\after\rho)~Q) \\
= & \eval (f\after\rho) (P \; \M{[]} Q).
\end{calc}

The proofs for the  binary operators nondeterministic choice (\CSPM{_ \|~\| _}),
sliding choice (\CSPM{_ [> _}), interrupt (\CSPM{_ /\\ _}), exception
(\CSPM{_ [\| _ \|> _}), and sequential composition (\CSPM{_ ; _})
are very similar.

For hiding, note that if $A$ is $T$-invariant, then
\begin{eqnarray*}
f(P \hide A) & = & f(P) \hide f(A).
\end{eqnarray*}
%
One the right-hand side, an event $e$ of~$P$ is hidden if $f(e) \in f(A)$; but
then $e \in A$, by Lemma~\ref{lem:T-invariant-inclusion}.  The rest of the
proof is then as for external choice.

\framebox{Guards}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Renaming}

The semantics of a renaming is a set of pairs of events $(e,e')$ indicating
that~$e$ is renamed to~$e'$.  \framebox{Or channels?}  I think the basic
definition should be extended to complete all events, and also add the
identity relation. 

We show the renaming itself satisfies the desired result.  Below, we use this
to show that the application of a renaming to a process also satisfies the
desired result. 
%
\begin{calc}
& f(\eval \rho~
   \rename{ e_1 \leftarrow e_1', \ldots, e_k \leftarrow e_k' \|  stmts}) \\
= & f \set{ (\eval \rho'~e_i, \eval \rho'~e_i') \|
     \rho' \in \evalStmts \rho~stmts, i \in \set{1,\ldots,k} } \\
= & \set{ (f(\eval \rho'~e_i), f(\eval \rho'~e_i')) \|
     \rho' \in \evalStmts \rho~stmts, i \in \set{1,\ldots,k} } \\
= & \com{inductive hypothesis} \\
  & \set{ (\eval (f\after\rho')~e_i, \eval (f\after\rho')~e_i') \|
     \rho' \in \evalStmts \rho~stmts, i \in \set{1,\ldots,k} } \\
= & \com{letting $\rho'' = f \after \rho'$} \\ 
  & \set{ (\eval \rho''~e_i, \eval \rho''~e_i') \|
     \rho'' \in f(\evalStmts \rho~stmts), i \in \set{1,\ldots,k} } \\
= & \com{inductive hypothesis, part~\ref{item:evalStmts}} \\
  & \set{ (\eval \rho''~e_i, \eval \rho''~e_i') \|
     \rho'' \in \evalStmts (f\after\rho)~stmts, i \in \set{1,\ldots,k} } \\
= & \eval (f\after\rho)
       \rename{ e_1 \leftarrow e_1', \ldots, e_k \leftarrow e_k' \|  stmts}.
\end{calc}


%% We need the following property for each relation~$R$ corresponding to a
%% renaming:
%% \begin{eqnarray}
%% \label{eqn:renaming-invariant}
%% (v,v') \in R \land f(v) = f(w) & \implies &
%%   \exists w' \spot (w,w') \in R \land f(v') = f(w').
%% \end{eqnarray}

Recall that each renaming satisfies condition~\ref{item:di-renaming} of
Definition~\ref{defn:data-independent}.  We will need the following lemma,
which shows that the value of a renaming is consistent with~$f$.
%
\begin{lemma}
\label{lem:renaming-invariant}
Consider a renaming $r = \rename{ e_1 \leftarrow e_1', \ldots, e_k
  \leftarrow e_k' \| stmts}$, and let $R = \eval \rho~ r$.  Then
%
\begin{eqnarray*}
%\label{eqn:renaming-invariant}
(v,v') \in R \land f(v) = f(w) & \implies &
  \exists w' \spot (w,w') \in R \land f(v') = f(w').
\end{eqnarray*}
\end{lemma}

%%%%%

\begin{proof}
Suppose $(v,v') \in R$ and $f(v) = f(w)$.  Suppose $(v,v')$ is produced by the
individual renaming $e_j \leftarrow e_j'$.  Consider the fields
$v_1,\ldots,v_k$ in~$v$ of type~$t$.  By condition~\ref{item:di-renaming} of
Definition~\ref{defn:data-independent}, these correspond to distinct
identifiers $x_1,\ldots,x_k$ in~$e_j$, each of which is bound by a generator
$x_i \,\M{<-}\, t$ in $stmts$.  Let the corresponding fields in~$w$ be
$w_1,\ldots,w_k$, so $f(v_i) = f(w_i)$ for each~$i$.
%
Consider the case where each~$x_i$ is bound to $w_i$, and every other
generator is bound as in the production of $(v,v')$.  These identifiers do not
appear in predicates in $stmts$, so each such predicate will still be true
with this binding.  Then the individual renaming $e_j \leftarrow e_j'$
produces $(w,w')$ for some $w'$ such that $f(v') = f(w')$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\framebox{** Partial events?}  Extend to total. 

The following lemma concerns semantic renaming, i.e.~the effect of applying a
renaming relation~$R$ to an LTS~$P$, denoted $P\rename{R}$.  
%
\begin{lemma}
\label{lem:renaming-f}
Let $P$ be an LTS, and $R$ a renaming relation.  Then
%
\begin{eqnarray*}
f(P\rename{R}) & = & (f(P))\rename{f(R)}.
\end{eqnarray*}
%
\end{lemma}

%%%%%

\begin{proof}
On each side, each state $Q$ of $P$ is replaced by $f(Q)$.
%
Suppose that $P$ contains a transition $Q \trans{e} Q'$.  Then the left-hand
side contains a transition $f(Q) \trans{f(e')} f(Q')$ for every $(e,e') \in
R$.  And the right-hand side contains a transition $f(Q) \trans{e''} f(Q')$
for every $e''$ such that $(f(e), e'') \in f(R)$.  We show that these
transitions correspond. 
%
\begin{itemize}
\item Consider such a transition of the left-hand side.  Then $(e,e') \in R$
  implies $(f(e),f(e') \in f(R)$, so the right-hand side has a corresponding
  transition.

\item Consider such a transition of the right-hand side.  If $(f(e), e'') \in
  f(R)$ then there is some $(e_1,e_2) \in R$ such that $f(e) = f(e_1)$ and
  $e'' = f(e_2)$.  But then Lemma~\ref{lem:renaming-invariant} implies
  $(e,e_3) \in R$ for some $e_3$ such that $f(e_3) = f(e_2) = e''$.  Hence
  the left-hand side has a corresponding transition.
\end{itemize}
\end{proof}

Let $r = \rename{ e_1 \leftarrow e_1', \ldots, e_k \leftarrow e_k' \|
  stmts}$ be a renaming.  We show the desired result is satisfied for the
application of~$r$ to a process~$P$.
%
\begin{calc}
& f(\eval \rho ~ (P~r)) \\
= & f((\eval \rho~P) \rename{ \eval \rho~r }) \\
= & \com{Lemma \ref{lem:renaming-f}} \\
  &  (f(\eval \rho~P)) \rename{ f(\eval \rho~r) } \\
= & \com{inductive hypothesis; earlier result} \\
  &  (\eval (f\after\rho)~P)) \rename{ \eval (f\after\rho)~r } \\
= & \eval (f\after\rho) (P~r).
\end{calc}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Binary Parallel operators.}

We write $\_ \parallel[\_] \_$ for a semantic operator that forms the parallel
composition of LTSs.  Recall that we assume that the identifiers bound by the
two components of the parallel composition are disjoint; this means that we
can restrict a ``global'' environment to the identifiers of each component, or
form the union of the environments of the two components.  For example, if
$(P, \rho_P) \trans{e} (P', \rho_P')$ and $(Q, \rho_Q) \trans{e} (Q',
\rho_Q')$ with $e \in A$, and $\rho_P$ and~$\rho_Q$ agree on common
identifiers, then $(P \parallel[A] Q, \rho_P \union \rho_Q) \trans{e} (P'
\parallel[A] Q', \rho_P' \union \rho_Q')$: the resulting environments
$\rho_P'$ and $\rho_Q'$ agree on common identifiers, so their union makes
sense.


We show
\begin{eqnarray}
\label{eqn:parallel-dist}
f((P,\rho_P) \parallel[A] (Q,\rho_Q)) & = & 
  f(P,\rho_P) \parallel[f(A)] f(Q,\rho_Q).
\end{eqnarray}
%
The tricky case is as follows.  Consider a transition of the right-hand side
corresponding to a synchronisation on event~$e \in f(A)$.  This must
correspond to transitions 
\[
(P,\rho_P) \trans{e_P} (P',\rho_P') \quad\mbox{and}\quad
(Q,\rho_Q) \trans{e_Q} (Q',\rho_Q') \quad
\mbox{with}\quad f(e_P) = f(e_Q) = e.
\]
The subsequent state is 
\[
f(P',\rho_P') \parallel[f(A)] f(Q',\rho_Q') = 
  (P',f\after\rho_P') \parallel[f(A)] (Q',f\after\rho_Q').
\]
If $e_P \ne e_Q$, then consider the first field in which they differ; say $e_p
= c.\vec{u}.v_P.\vec{w_P}$ and $e_Q = c.\vec{u}.v_Q.\vec{w_Q}$ with $v_P \ne
v_Q$ but $f(v_P) = f(v_Q)$.
%
Then by assumption \framebox{???}, one of the processes must be willing to
accept an arbitrary value in this position; without loss of generality,
suppose this is~$P$, so $P$ can also perform $e_P' = c.\vec{u}.v_Q.\vec{w_P}$.
If the transition binds $x$ to this field in $P$'s environment, then this
corresponds to the transition $(P,\rho_P) \trans{e_P'} (P,\rho_P'')$ where
$\rho_P'' = \rho_P' \oplus \set{x \mapsto v_Q}$.  Note that $f(e_P') = e$ and
$f\after\rho_P'' = f \after \rho_P'$. 
%
Continuing in this way, we obtain transitions
\[
\begin{array}{c}
(P,\rho_P) \trans{e'} (P',\hat\rho_P) \quad\mbox{and}\quad
(Q,\rho_Q) \trans{e'} (Q',\hat\rho_Q) \\
\mbox{with}\quad f(e')  = e, \quad 
f \after \hat\rho_P = f \after \rho_P', \quad
f \after \hat\rho_Q = f \after \rho_Q'.
\end{array}
\]
Also, note that $f(e') = e \in f(A)$, so $e' \in A$, by
Lemma~\ref{lem:T-invariant-inclusion}, since $A$ is $T$-invariant.
%
Hence, in the left-hand side of (\ref{eqn:parallel-dist}), the two
components can synchronise on~$e'$, giving a transition on $f(e') = e$, and
producing state 
\[
f( (P',\hat\rho_P) \parallel[A] (Q',\hat\rho_Q) ) .
\]
This matches the original transition of the right-hand side.

Other cases ---that every synchronisation of the left-hand side is matched by
a synchronisation of the right-hand side, and that non-synchronisation
transitions match--- can be proved in a similar but more straightforward way. 

\framebox{Other parallel ops}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Indexed operators.}

Below we write $\Extchoice$ for a semantic operator: ``$\Extchoice S$''
represents an external choice over the set of processes in~$S$, each of which
is an LTS\@.  We have
%
\begin{eqnarray*}
f(\Extchoice S) & = & \Extchoice \set{f(P) \| P \in S}.
\end{eqnarray*}
%
For every $P \in S$, and for every transition $P \trans{e} Q$, the initial
state of each of the above processes has a transition labelled with $f(e)$ to
$f(Q)$.

We then have the following (for syntactic expressions~$S$ and~$P$):
%
\begin{calc}
& f(\eval \rho ~ (\M{**[]} pat \leftarrow S \spot P)) \\
= & f(\Extchoice \set{\eval (\bind \rho ~ pat~ v) ~ P \| 
        v \in \eval \rho~ S, \matches \rho ~ pat ~ v}) \\
= & \com{above result} \\
& \Extchoice \set{f(\eval (\bind \rho ~ pat~ v) ~ P) \| 
        v \in \eval \rho~ S, \matches \rho ~ pat ~ v} \\
= & \com{inductive hypothesis for~$P$} \\
& \Extchoice \set{\eval (f \after \bind \rho ~ pat~ v) ~ P) \| 
        v \in \eval \rho~ S, \matches \rho ~ pat ~ v} \\
= & \com{inductive hypothesis, parts~\ref{item:matches} and~\ref{item:bind}} \\
& \Extchoice 
  \begin{align}
  \set{\eval (\bind (f \after \rho) ~ pat~ (f(v))) ~ P) \| \\
  \qquad   v \in \eval \rho~ S, \matches (f\after\rho) ~ pat ~ (f(v)) }
  \end{align} \\
= & \com{letting $w = f(v)$} \\
& \Extchoice 
  \set{\eval (\bind (f \after \rho) ~ pat~ w) ~ P) \| 
     w \in f(\eval \rho~ S), \matches (f\after\rho) ~ pat ~ w } \\
= & \com{inductive hypothesis for~$S$} \\
& \Extchoice 
  \set{\eval (\bind (f \after \rho) ~ pat~ w) ~ P) \| 
     w \in \eval (f\after\rho)~ S, \matches (f\after\rho) ~ pat ~ w } \\
= & \eval (f\after\rho) ~ (\M{**[]} pat \leftarrow S \spot P).
\end{calc}

